---
title: "Benchmark mzML vs mzPeak"
format: html
---

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(microbenchmark)
devtools::load_all("/home/gabriele/Desktop/EURAC/MetaRbolomics4Galaxy/MsBackendMzPeak/")
baseDir <- "/home/gabriele/Desktop/EURAC/MetaRbolomics4Galaxy/MsBackendMzPeak_test/"
```

# Comparison between mzML and mzPeak

In this document I will benchmark the functions to read *mzML* and *mzPeak*:
- `Spectra`+`MsBackendMzR` for *mzML*
- `MsBackendMzPeak` for *mzPeak*

# Small dataset

First, I test a smal dataset with 2 files:
- MsDataHub::PestMix1_DDA.mzML
- MsDataHub::PestMix1_SWATH.mzML

The files are converted to mzPeak with the convertor in
[HUPO-PSI/mzPeak](https://github.com/HUPO-PSI/mzPeak#conversion-program)

## Example convert files

To convert the `mzML` files to `mzPeak` use th function

```{r, eval = FALSE}
convertToMzPeak(path = paste0(baseDir, "MTBLS1129"),
                pattern = ".mzML$",
                output_dir = paste0(baseDir, "MTBLS1129"))
```

## Benchmark mzML

Metadata

```{r}
small_mzml <- c(MsDataHub::PestMix1_DDA.mzML(), MsDataHub::PestMix1_SWATH.mzML())
sps_sciex <- Spectra(small_mzml, source = MsBackendMzR())

microbenchmark({
    mtd_small_mzml <- spectraData(sps_sciex)
}, times = 5L, unit = "s")

```

Peaks single spectra

```{r}
microbenchmark({
    peaks_small_mzml <- peaksData(sps_sciex, scan = 1L)
}, times = 5L, unit = "s")

```

Peaks single spectra

```{r}
microbenchmark({
    peaks_small_mzml <- peaksData(sps_sciex, scan = 1:100)
}, times = 5L, unit = "s")

```

Peaks

```{r}
microbenchmark({
    peaks_small_mzml <- peaksData(sps_sciex)
}, times = 5L, unit = "s")

```

## Benchmark mzPeak

Load only spectrum index of the mzPeak archive

```{r}
small_mzPeak <- list.files(paste0(baseDir, "small_data_files"),
                            pattern = ".mzPeak", full.names = TRUE)

microbenchmark({
    index_small_mzpeak <- load_mzpeak(small_mzPeak)
}, times = 5L, unit = "s")

```

Read minimal metadata of all Spectra of the mzPeak based on iterator.

- `spectrum_metadata()` is the original function,
- `spectrum_metadata_v2()` is the new function based on Joshua suggestion to use `load_all_spectrum_metadata()`

```{r}
small_mzPeak <- list.files(paste0(baseDir, "small_data_files"),
                            pattern = ".mzPeak", full.names = TRUE)

microbenchmark({
    desc_min_small_mzpeak_min <- spectrum_metadata(small_mzPeak, minimal = TRUE)
},{
    desc_min_small_mzpeak_min_v2 <- spectrum_metadata_v2(small_mzPeak, minimal = TRUE)
}, times = 5L, unit = "s")
```

Read the metadata of all Spectra of the mzPeak based on iterator

```{r}
microbenchmark({
    desc_small_mzpeak <- spectrum_metadata(small_mzPeak, minimal = FALSE)
},{
    desc_small_mzpeak_v2 <- spectrum_metadata_v2(small_mzPeak, minimal = FALSE)
}, times = 5L, unit = "s")

```


Read the Centroid spectrum of 1 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r}
microbenchmark({
    peaks_small_mzpeak_filt_1 <- lapply(small_mzPeak, spectrum_peaks_by_id, index = 1L)
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of 100 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r}
microbenchmark({
    peaks_small_mzpeak_filt_10 <- lapply(small_mzPeak, spectrum_peaks_by_id, index = 1:100)
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of all Speactra of the mzPeak (`Spectra_peaks.parquet`).
One single file

- `spectrum_peaks()` is the original function using `get_spectrum()`, multitreaded on files in RUST,
- `spectrum_peaks_v2()` is the new function based on Joshua suggestion to use `query_peaks()`, parse the results and convert, multitreaded on files in RUST,
- `spectrum_peaks_v3()` is the new function based on Joshua suggestion to use `query_peaks()`, `arrow_extendr` to convert the `RecordBatch` to `Robj` in Rust and stream it to R via `nanoarrow::convert_array_stream()`, single thread.

```{r}
microbenchmark({
    peaks_small_mzpeak <- spectrum_peaks(small_mzPeak[[1]])
},{
    peaks_small_mzpeak <- spectrum_peaks_v2(small_mzPeak[[1]])
},{
    peaks_small_mzpeak <- spectrum_peaks_v3(small_mzPeak[[1]])
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of all Speactra of the mzPeak (`Spectra_peaks.parquet`)

```{r}
microbenchmark({
    peaks_small_mzpeak <- spectrum_peaks(small_mzPeak)
},{
    peaks_small_mzpeak <- spectrum_peaks_v2(small_mzPeak)
},{
    peaks_small_mzpeak <- spectrum_peaks_v3(small_mzPeak)
}, times = 5L, unit = "s")

```

# Mid-size dataset

I test with an example dataset from mzPeak based on the Metabolights dataset
[MTBLS1129](https://object.storage.eu01.onstackit.cloud/v09/sdrf.html)

The files are converted to mzPeak with the convertor in
[HUPO-PSI/mzPeak](https://github.com/HUPO-PSI/mzPeak#conversion-program)

## Benchmark mzML

Metadata

```{r}
mid_mzml <- list.files(paste0(baseDir, "MTBLS1129"), pattern = "mzML",
                        full.names = TRUE)
microbenchmark({
    sps_sciex <- Spectra(mid_mzml, source = MsBackendMzR())
    mtd_mid_mzml <- spectraData(sps_sciex)
}, times = 5L, unit = "s")

```

Peaks

```{r}
microbenchmark({
    sps_sciex <- Spectra(mid_mzml, source = MsBackendMzR())
    peaks_mid_mzml <- peaksData(sps_sciex)
}, times = 5L, unit = "s")

```

## Benchmark mzPeak

Load only spectrum index of the mzPeak archive

```{r}
mid_mzPeak <- list.files(paste0(baseDir, "MTBLS1129/convert_v20260716"),
                            pattern = ".mzPeak", full.names = TRUE)

microbenchmark({
    index_mid_mzpeak <- load_mzpeak(mid_mzPeak)
}, times = 5L, unit = "s")

```

Read minimal metadata of all Spectra of the mzPeak based on iterator

```{r}
microbenchmark({
    desc_min_mid_mzpeak <- spectrum_metadata(mid_mzPeak, minimal = TRUE)
},{
    desc_min_mid_mzpeak_v2 <- spectrum_metadata_v2(mid_mzPeak, minimal = TRUE)
}, times = 5L, unit = "s")

```

Read the metadata of all Spectra of the mzPeak based on iterator

```{r}
microbenchmark({
    desc_mid_mzpeak <- spectrum_metadata(mid_mzPeak, minimal = FALSE)
},{
    desc_mid_mzpeak_v2 <- spectrum_metadata_v2(mid_mzPeak, minimal = FALSE)
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of 1 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r}
microbenchmark({
    peaks_mid_mzpeak_filt_1 <- lapply(mid_mzPeak, spectrum_peaks_by_id, index = 1L)
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of 100 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r}
microbenchmark({
    peaks_mid_mzpeak_filt_10 <- lapply(mid_mzPeak, spectrum_peaks_by_id, index = 1:100)
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of all Speactra of the mzPeak (`Spectra_peaks.parquet`)
Read 1 single file

```{r}
microbenchmark({
    peaks_mid_mzpeak <- spectrum_peaks(mid_mzPeak[[1]])
}, {
    peaks_mid_mzpeak <- spectrum_peaks_v2(mid_mzPeak[[1]])
},{
    peaks_mid_mzpeak <- spectrum_peaks_v3(mid_mzPeak[[1]])
}, times = 5L, unit = "s")

```

Read the Centroid spectrum of all Speactra of the mzPeak (`Spectra_peaks.parquet`)
Read multithread with iterator

```{r}
microbenchmark({
    peaks_mid_mzpeak <- spectrum_peaks(mid_mzPeak)
}, {
    peaks_mid_mzpeak <- spectrum_peaks_v2(mid_mzPeak)
},{
    peaks_mid_mzpeak <- spectrum_peaks_v3(mid_mzPeak)
}, times = 5L, unit = "s")

```


# MidBig-size dataset

I test with an example dataset from mzPeak based on the MassIVE dataset
[MSV000090156](https://massive.ucsd.edu/ProteoSAFe/dataset.jsp?task=06bd49807caa4390961fb827606a8696)
with 169 files.

The files are converted to mzPeak with the convertor in
[HUPO-PSI/mzPeak](https://github.com/HUPO-PSI/mzPeak#conversion-program)

## Benchmark mzML

Metadata

```{r, eval = FALSE}
midbig_mzml <- list.files(paste0(baseDir, "big_data_files"), pattern = "mzML",
                        full.names = TRUE)
microbenchmark({
    sps_sciex <- Spectra(midbig_mzml, source = MsBackendMzR())
    mtd_midbig_mzml <- spectraData(sps_sciex)
}, times = 1L, unit = "s")

```

Peaks

```{r, eval = FALSE}
microbenchmark({
    sps_sciex <- Spectra(midbig_mzml, source = MsBackendMzR())
    peaks_midbig_mzml <- peaksData(sps_sciex)
}, times = 1L, unit = "s")

```

## Benchmark mzPeak

Load only spectrum index of the mzPeak archive

```{r, eval = FALSE}
midbig_mzPeak <- list.files(paste0(baseDir, "big_data_files"),
                            pattern = ".mzPeak", full.names = TRUE)

microbenchmark({
    index_midbig_mzpeak <- load_mzpeak(midbig_mzPeak)
}, times = 1L, unit = "s")

```

Read minimal metadata of all Spectra of the mzPeak based on iterator

```{r, eval = FALSE}
microbenchmark({
    desc_min_midbig_mzpeak <- spectrum_metadata_v2(midbig_mzPeak, minimal = TRUE)
}, times = 1L, unit = "s")

```

Read the metadata of all Spectra of the mzPeak based on iterator

```{r, eval = FALSE}
microbenchmark({
    desc_midbig_mzpeak_iter <- spectrum_metadata_v2(midbig_mzPeak, minimal = FALSE)
}, times = 1L, unit = "s")

```

Read the Centroid spectrum of 1 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r, eval = FALSE}
microbenchmark({
    peaks_midbig_mzpeak_filt_1 <- lapply(midbig_mzPeak, spectrum_peaks_by_id, index = 1L)
}, times = 1L, unit = "s")

```

Read the Centroid spectrum of 100 Speactra of the mzPeak (`Spectra_peaks.parquet`)
using `get_spectra_batch()`

```{r, eval = FALSE}
microbenchmark({
    peaks_midbig_mzpeak_filt_10 <- lapply(midbig_mzPeak, spectrum_peaks_by_id, index = 1:100)
}, times = 1L, unit = "s")

```

Read the Centroid spectrum of all Speactra of the mzPeak (`Spectra_peaks.parquet`)
Read multithread with iterator

```{r, eval = FALSE}
microbenchmark({
    peaks_midbig_mzpeak <- spectrum_peaks(midbig_mzPeak)
}, times = 1L, unit = "s")

```
